Uniformly Ergodic Data-Augmented MCMC for Fitting Stochastic Epidemic Models to Incidence Data

### 2021-10-15 Manuscripts/Article/Article.tex
test overleaf
move .bib file
Updates from Overleaf

	The following characteristics of our DA-MCMC algorithm are worth noting.
	First, initializing the Markov chain only requires setting $\theta^{(0)} = (\beta^{(0)}, \gamma^{(0)})$ since $\Z^{(0)}$ can be generated from the PD-SIR process conditionally on $\theta^{(0)}$ and $\Y$ alone.
	Second, since the PD-SIR closely approximates the SIR model, the acceptance rate in the Metropolis-Hastings step for the latent data is relatively high. For large population, however, the acceptance rate may be too low, thereby hindering the mixing of the Markov chain. To address this issue, we propose to update the event times of only a fraction $0<\rho\le 1$ of the individuals in the population. Smaller values for $\rho$ result in smaller jumps in the latent space and a larger acceptance rate.
	Third, if all event times are updated ($\rho = 1$), then the proposed and current latent data are independent conditionally on the current values of the parameters: $\Z^{(k-1)} \perp \Z^* | \theta^{(k)}$. This contrasts with existing DA-MCMC algorithms which only update a small fraction of the latent data per iteration.
	%Fourth, since the latent data are generated from a process that only resembles the SIR model, we ensure that the Markov chain converges to the correct distribution $\pi(\theta, \Z|\Y)$ by proposing and accepting the latent data according to a Metropolis-Hastings scheme. This results in exact Bayesian inference targeting the posterior distribution under the original SIR model.
	%Fifth, 
	Fourth, not only is proposing latent data from the RD-SIR process extremely fast since all random variables can be generated via the inverse-CDF method, but ensuring that these data are compatible with the observed data can be done at no additional cost, making the proposal scalable to large outbreaks.
	Finally, we show in the following section that the algorithm generates a Markov chain that is \textit{uniformly ergodic}, the strongest form of ergodicity for a Markov chain \cite{Tierney.1994}. %The proof of this result is provided in Appendix \ref{app:uni}.
	% our DA-MCMC does only requires I(0), not S(0).

@@

	In this section, the convergence properties of the DA-MCMC algorithm are examined with a simulation study. Each data set in this study is simulated from the stochastic SIR process and the prior distributions on the parameters are the independent weakly informative distributions $\beta \sim Ga(0.1, 1)$ and $\gamma \sim Ga(1,1)$.

	First, the mixing property of the Markov chain is assessed in a medium-size population whose initial configuration is $(S(0), I(0)) = (1000, 10)$. The parameters are $(\beta, \gamma) = (0.003, 1)$ and the process is observed until time $t_{end} = 6$ when the pandemic has completed most of its course but is not over yet (see Figure \ref{fig:long}, where $I(t_{end}) = 55$). The numbers of infections in $K = 10$ time intervals of equal length are observed and correspond to $T_{1:K} = (40, 111, 193, 259, 178, 93, 29, 19, 9, 6)$.
	The Markov chain is run for $100,000$ iterations from the initial values $(\beta^{(1)}, \gamma^{(1)}) = (0.0003, 0.1) = (\beta/10, \gamma/10)$ and the event times of $202$ ($\rho = 0.2$) individuals are updated in each Metropolis-Hastings step for the latent data. The algorithm takes in $2.5$ minutes to complete on a personal laptop. 
	Figure \ref{fig:traceplot} shows the traceplots of the parameters $\beta$ and $\gamma$, the log likelihood and the basic reproduction number $R_0 = S(0) \beta / \gamma$, which corresponds to the average number of secondary infections generated by an infectious individual in a susceptible population. We observe that the Markov chain quickly migrates from the low density region of the space where $(\beta^{(1)}, \gamma^{(1)})$ is located to a high density region where it mixes well. The acceptance rate of the latent spaces proposed in the Metropolis-Hastings step is $0.11$.
	Excluding the first $10,000$ iterations of the Markov chain as a burn-in, the posterior means of $\beta$, $\gamma$ and $R_0$ are respectively $0.00304$, $0.995$ and $3.07$.

@@

	\begin{figure}
		\centering
		\includegraphics[scale = 0.3]{E1_trajectories}
		\caption{Trajectories of the $S$, $I$ and $R$ compartments of a SIR process for a medium-size population ($S_0 = 1,000$).}
		\label{fig:long}
	\end{figure}

	\begin{figure}
		\centering
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_no_burn_beta_tp}
			\caption{$\beta$}
			\label{fig:traceplot_beta}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_burn_beta_acf}
			\caption{$\gamma$}
			\label{fig:traceplot_gamma}
		\end{subfigure}
		\\
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_no_burn_R0_hist}
			\caption{Log likelihood}
			\label{fig:traceplot_loglik}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_burn_R0_acf}
			\caption{$R_0$}
			\label{fig:traceplot_R0}
		\end{subfigure}
		\caption{Traceplots of the Markov chain (including burn-in).}
		\label{fig:traceplot}
	\end{figure}

@@

	\bibliographystyle{plain}
	\bibliography{../bibliography}

>>>>

	The following characteristics of our DA-MCMC algorithm are worth noting.
	First, initializing the Markov chain only requires setting $\theta^{(0)} = (\beta^{(0)}, \gamma^{(0)})$ since $\Z^{(0)}$ can be generated from the PD-SIR process conditionally on $\theta^{(0)}$ and $\Y$ alone.
	Second, since the PD-SIR closely approximates the SIR model, the acceptance rate in the Metropolis-Hastings step for the latent data is relatively high. For large population, however, the acceptance rate may be too low, thereby hindering the mixing of the Markov chain. To address this issue, we propose to update the event times of only a fraction $0<\rho\le 1$ of the individuals in the population. Smaller values for $\rho$ result in smaller jumps in the latent space and a larger acceptance rate.
	Third, if all event times are updated ($\rho = 1$), then the proposed and current latent data are independent conditionally on the current values of the parameters: $\Z^{(k-1)} \perp \Z^* | \theta^{(k)}$. This contrasts with existing DA-MCMC algorithms which only update a small fraction of the latent data per iteration.
	%Fourth, since the latent data are generated from a process that only resembles the SIR model, we ensure that the Markov chain converges to the correct distribution $\pi(\theta, \Z|\Y)$ by proposing and accepting the latent data according to a Metropolis-Hastings scheme. This results in exact Bayesian inference targeting the posterior distribution under the original SIR model.
	%Fifth, 
	Fourth, not only is proposing latent data from the RD-SIR process extremely fast since all random variables can be generated via the inverse-CDF method, but ensuring that these data are compatible with the observed data can be done at no additional cost, making the proposal scalable to large outbreaks.
	Finally, we show in the following section that the algorithm generates a Markov chain that is \textit{uniformly ergodic}, the strongest form of ergodicity for a Markov chain \cite{Tierney.1994}. %The proof of this result is provided in Appendix \ref{app:uni}.
	% our DA-MCMC only requires I(0), not S(0).

@@

	We examine the convergence properties of the proposed DA-MCMC algorithm with simulation studies. First, we assess the performance of the algorithm in a medium-sized population of with over $1000$ individuals. We simulate a pandemic with parameters $(\beta, \gamma) = (0.0025, 1)$, which gives $R_0 = 2.5$, starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$ when the pandemic has completed most of its course but is not over yet (see Figure \ref{fig:fig:E1_trajectories}, where $I(t_{end}) = 55$). The numbers of infections in $K = 10$ time intervals of equal length are observed, corresponding to $I_{1:K} = (40, 111, 193, 259, 178, 93, 29, 19, 9, 6)$.	
	We set the independent weakly informative prior distributions on the parameters $\beta \sim Ga(0.01, 1)$ and $\gamma \sim Ga(2,2)$.
	The Markov chain is initialized in a low density region at $(\beta^{(0)}, \gamma^{(0)}) = (0.0003, 0.1) = (\beta/10, \gamma/10)$ and is run for $1$ million iterations. We keep every tenth draw for memory reason and set $\rho = 0.1$ so that the trajectories of $\rho (S(0)+I(0)) = 101$ individuals are updated in each Metropolis-Hastings step. The algorithm runs takes $25$ minutes to complete on a personal laptop.

	Figure \ref{fig:traceplot} shows the traceplots of the parameters $\beta$ and $\gamma$, the log likelihood and the basic reproduction number $R_0 = S(0) \beta / \gamma$, which corresponds to the average number of secondary infections generated by an infectious individual in a susceptible population. We observe that the Markov chain quickly migrates from the low density region of the space where $(\beta^{(1)}, \gamma^{(1)})$ is located to a high density region where it mixes well. The acceptance rate of the latent spaces proposed in the Metropolis-Hastings step is $0.11$.
	Excluding the first $10,000$ iterations of the Markov chain as a burn-in, the posterior means of $\beta$, $\gamma$ and $R_0$ are respectively $0.00304$, $0.995$ and $3.07$.

@@

	\begin{figure}
		\centering
		\begin{subfigure}[b]{0.42\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_trajectories}
			\caption{Compartment trajectories}
			\label{fig:E1_trajectories}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_no_burn_gamma_tp}
			\caption{$\gamma$}
			\label{fig:traceplot_gamma}
		\end{subfigure}
		\\
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_no_burn_gamma_hist}
			\caption{Log likelihood}
			\label{fig:traceplot_loglik}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_burn_gamma_acf}
			\caption{$R_0$}
			\label{fig:traceplot_R0}
		\end{subfigure}
		\caption{Performance of the proposed DA-MCMC in a medium-sized population $(S(0), I(0)) = (1000, 10)$.}
		\label{fig:E1}
	\end{figure}

@@

	\bibliographystyle{plain}
	\bibliography{bibliography}
